Model Finding for Recursive Functions in SMT

### 2015-04-29 tex/main.tex
Minor

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\M}{\mathcal{M}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{W}}

@@

We write $\terms( t )$ to denote the set of subterms in $t$.
We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

@@

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universal quantification in the left disjunct,
state-of-the-art SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of sort $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{equation*}
The universal quantification in the resulting formula is over an uninterpreted sort $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
$\square$
\end{example}

@@

Now, examining the second encoding (the latter nine columns), 
we see that using the translation $\conv$ drastically improves the ability of configurations for determining satisfiability.
In total, {\bf cvc4+f} finds 331 of the 885 benchmarks to be satisfiable, including 9 benchmarks in the original {\bf leon} benchmark set.
The performance of {\bf z3} for counterexamples also improves drastically, as it solves 112 satisfiable,
including 10 that were unsolved by {\bf cvc4+f}.
Moreover, the translation $\conv$ helps \ziii for unsatisfiable responses as well,
as we see \ziii solves a total of 209 whereas it solves only 187 before the translation.
This number is 2 fewer than {\bf cvc4}, which solves 211 unsatisfiable.
Thus, in most cases the translation $\conv$ does not seriously degrade performance for unsatisfiable benchmark,
and in some cases may actually aid the solver for determining unsatisfiability
\footnote{The exception here is {\bf cvc4+i}, which makes use of type information of quantified formulas to make inductive arugments.}.

>>>>

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\M}{\mathcal{M}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{W}}

@@

We write $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

@@

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universal quantification in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of sort $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{equation*}
The universal quantification in the resulting formula is over an uninterpreted sort $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
$\square$
\end{example}

@@

Now, examining the second encoding (the latter nine columns), 
we see that using the translation $\conv$ drastically improves the ability of configurations for answering satisfiable.
In total, {\bf cvc4+f} finds 331 of the 885 benchmarks to be satisfiable, including 9 benchmarks in the original {\bf leon} benchmark set.
The performance of {\bf z3} for counterexamples also improves drastically, as it solves 112 satisfiable,
including 10 that were unsolved by {\bf cvc4+f}.
Moreover, the translation $\conv$ helps \ziii for unsatisfiable responses as well,
as we see \ziii solves a total of 209 whereas it solves only 187 before the translation.
This number is 2 fewer than {\bf cvc4}, which solves 211 unsatisfiable.
Thus, in most cases the translation $\conv$ does not seriously degrade performance for unsatisfiable benchmark,
and in some cases may actually aid the solver for determining unsatisfiability
\footnote{The exception here is {\bf cvc4+i}, which makes use of type information of quantified formulas to make inductive arugments.}.
